### 2021-06-01 chaps/thesis/chap1/chapitre1.tex
🚚 rename thesis directory
📝 add a little intro

\boitemagique{TODO}{Indiquer qu'il s'agit d'un copier-coller de l'article \cite{Crouseilles:2019b}, avec peut-être une petite intro et conclusion pour ajouter un peu plus d'info sur l'algorithme d'obtention de CFL etc.}
\boitemagique{TODO}{Revoir les références bibliographiques de ce chapitre (les ajouter dans mon \texttt{biblio.bib}}

Merci Lukas et Nicolas pour la rédaction de ce chapitre.

\selectlanguage{english}

%% article
\input{\localPath/sections/article}

\selectlanguage{french}

>>>>

%\boitemagique{TODO}{Indiquer qu'il s'agit d'un copier-coller de l'article~\cite{Crouseilles:2019b}, avec peut-être une petite intro et conclusion pour ajouter un peu plus d'info sur l'algorithme d'obtention de CFL etc.}

Ce chapitre a pour objectif de présenter les outils de calcul de stabilité, les schémas exponentielles, et se familiariser avec des méthodes d'ordre élevé comme le schéma WENO5. Le calcul de stabilité pour une méthode WENO a été initié dans~\cite{Wang:2007} et~\cite{Motamed:2010}, le calcul de condition de CFL a été poursuivit dans~\cite{Lunet:2017} ; ce chapitre continue ce travail et propose un algorithme d'estimation de condition de CFL entre un schéma WENO et une méthode Runge-Kutta donnée, en plus d'étudier la stabilité des méthodes exponentielles en temps pour une équation de transport, avec une application a des équations cinétiques que sont les équations de la physique des plasmas. Ce travail est synthétisé par cet article qui est un travail colaboratif avec Nicolas Crouseilles\footnote{Univ Rennes, Inria Bretagne Atlantique (MINGuS) \& ENS Rennes} et Lukas Einkemmer\footnote{Department of Mathematics, Univeristy of Innsbruck, Austria} qui a mené a un article \emph{Exponential methods for solving hyperbolic problems with application to kinetic equations} publié dans \emph{Journal of Computational Physics} en 2020~\cite{Crouseilles:2019b}.

\selectlanguage{english}
%% article
\input{\localPath/sections/article}
\selectlanguage{french}

### 2021-06-01 chaps/thesis/chap1/sections/article.tex
📝 add a little intro

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $y_{\max}=\min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t)$. As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that : $i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

>>>>

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $$y_{\max}=\min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t).$$ As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that : $$i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D},$$ where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}
